---
title: "Exam 3 Skills Test"
author: "Kate Fox"
date: "April 8, 2026"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(tidyverse)
library(broom)
```

# Load & Clean Faculty Salaries
In this section, I loaded the faculty salary dataset and transformed it into a tidy format. Tidying the data ensures that each row represents a single observation of salary by state, rank, and tier, making it easier to analyze and visualize.
```{r}
# Load dataset
salaries <- read.csv("FacultySalaries_1995.csv")
glimpse(salaries)

# Clean the data set
salaries_tidy <- salaries %>%
  pivot_longer(
    cols = c(AvgFullProfSalary, AvgAssocProfSalary, AvgAssistProfSalary),
    names_to = "Rank",
    values_to = "Salary"
  ) %>%
  mutate(
    Rank = case_when(
      Rank == "AvgFullProfSalary" ~ "Full",
      Rank == "AvgAssocProfSalary" ~ "Associate",
      Rank == "AvgAssistProfSalary" ~ "Assistant"
    )
  )
```

To better understand salary distributions, I visualized salaries across faculty ranks and institutional tiers.
```{r}
salaries_tidy %>%
  filter(Tier %in% c("I", "IIA", "IIB")) %>%
  ggplot(aes(x = Rank, y = Salary, fill = Rank)) +
  geom_boxplot() +
  facet_wrap(~ Tier) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Rank", y = "Salary")
```

# ANOVA Model
To evaluate whether State, Tier, and Rank have significant effects on salary, I fit an ANOVA model without interaction terms. This model allows us to assess the independent contribution of each factor.
```{r}
anova_model <- aov(Salary ~ State + Tier + Rank, data = salaries_tidy)
summary(anova_model)
```

The ANOVA output identifies which factors significantly influence salary variation across institutions.

# Load & Tidy Juniper Oils
Next, I loaded the Juniper Oils dataset and reshaped it into a tidy format. Each row now represents a single chemical concentration measurement for a given observation.
```{r}
# Load dataset
oils <- read_csv("Juniper_Oils.csv")

# List of chemical variables
chemicals <- c("alpha-pinene","para-cymene","alpha-terpineol","cedr-9-ene",
               "alpha-cedrene","beta-cedrene","cis-thujopsene","alpha-himachalene",
               "beta-chamigrene","cuparene","compound 1","alpha-chamigrene",
               "widdrol","cedrol","beta-acorenol","alpha-acorenol",
               "gamma-eudesmol","beta-eudesmol","alpha-eudesmol",
               "cedr-8-en-13-ol","cedr-8-en-15-ol","compound 2","thujopsenal")

# Tidy dataset
oils_tidy <- oils %>%
  pivot_longer(
    cols = all_of(chemicals),
    names_to = "ChemicalID",
    values_to = "Concentration"
  )
```

# Faceted Visualization
To explore how chemical concentrations vary with time since burn, I created faceted plots for each compound. This allows for easy comparison of trends across many chemicals.
```{r}
ggplot(oils_tidy, aes(x = YearsSinceBurn, y = Concentration)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~ChemicalID, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Chemical Concentration vs. Years Since Burn",
    x = "Years Since Burn",
    y = "Concentration"
  )
```

# Linear Model for Significant Chemicals
To determine which chemicals are significantly affected by time since burn, I fit a linear model including an interaction between chemical identity and time. This allows each chemical to have its own slope.
```{r}
glm_model <- lm(Concentration ~ ChemicalID * YearsSinceBurn, data = oils_tidy)

# Extract model results
results <- tidy(glm_model)

# Filter significant terms
significant_results <- results %>%
  filter(p.value < 0.05)

significant_results
```

The significant interaction terms indicate which chemicals show meaningful changes in concentration over time since burn.