---
title: 'Assignment 9: Graduate School Admissions'
author: "Kate Fox"
date: "2026-03-24"
output: html_document
---

### Introduction ###

In this assignment, I analyzed predictors of graduate school admissions using GPA, GRE, and undergraduate school ranking.

### Data Loading & Cleaning ###
```{r}
library(tidyverse)
admissions <- read_csv("GradSchool_Admissions.csv")
glimpse(admissions)
summary(admissions)

admissions <- admissions %>%
  mutate(
    admit = as.factor(admit),
    rank = as.factor(rank)
  )
``` 
### Exploratory Data Analysis ###
```{r}
# GPA vs Admission
ggplot(admissions, aes(x = gpa, fill = admit)) +
  geom_density(alpha = 0.5) +
  labs(title = "GPA Distribution by Admission Status", x = "GPA", fill = "Admit")

# GRE vs Admission
ggplot(admissions, aes(x = gre, fill = admit)) +
  geom_density(alpha = 0.5) +
  labs(title = "GRE Distribution by Admission Status", x = "GRE", fill = "Admit")

# Rank vs Admission
ggplot(admissions, aes(x = rank, fill = admit)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Admissions by Undergraduate Rank", y = "Proportion", fill = "Admit")
```
The first two graphs indicate that higher GPA and GRE increase admission probability. A higher rank (1) has higher proportion of admitted students, so rank does matter.

### Model Building ###
# Model 1
```{r}
# Model 1: GPA only
mod1 <- glm(admit ~ gpa, data = admissions, family = binomial)
summary(mod1)

# Model 2: GPA + GRE
mod2 <- glm(admit ~ gpa + gre, data = admissions, family = binomial)
summary(mod2)

# Model 3: GPA + GRE + Rank
mod3 <- glm(admit ~ gpa + gre + rank, data = admissions, family = binomial)
summary(mod3)

# Compare models
AIC(mod1, mod2, mod3)
```
GPA is significant on it's own. Adding GRE improves the model slightly. Rank is significant, so model 3 is the best when comparing with AIC.

### Model Evaluation & Predicted Probabilities ###
# Add Predicted Probabilities from Model 3
```{r}
admissions <- admissions %>%
  mutate(pred_prob = predict(mod3, type = "response"))
admissions
```
# Plot Predicted Probabilities by Actual Admission
```{r}
ggplot(admissions, aes(x = pred_prob, fill = admit)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) +
  labs(title = "Predicted Admission Probabilities (mod3)",
       x = "Predicted Probability of Admission",
       y = "Count",
       fill = "Admit")
```
The histogram shows that students who were admitted tend to have higher predicted
probabilities, which suggests that our model is reasonable in comparing admitted and
non-admitted students.

### Odds Ratios for Interpretation ###
```{r}
# Odds Ratios
odds_ratios <- exp(coef(mod3))
odds_ratios

# Display as a Table
data.frame(
  Predictor = names(odds_ratios),
  Odds_Ratio = odds_ratios
)
```
Odds ratios show how each predictor affects the odds of admission. A GPA odds ratio of >1 means higher GPA increases the odds of being admitted.

#Visualizing Effects of Rank
```{r}
### Admission Probability by Rank ###
# Predict admission probability for each rank at average GRE and GPA
avg_gre <- mean(admissions$gre)
avg_gpa <- mean(admissions$gpa)

rank_pred <- data.frame(
  gre = avg_gre,
  gpa = avg_gpa,
  rank = factor(1:4, levels = levels(admissions$rank))
)
rank_pred$pred_prob <- predict(mod3, newdata = rank_pred, type = "response")

ggplot(rank_pred, aes(x = rank, y = pred_prob)) +
  geom_col(fill = "steelblue") +
  labs(title = "Predicted Probability of Admission by Rank",
       x = "Undergraduate Institution Rank",
       y = "Predicted Probability")
```
Holding GRE and GPA constant, top-tier schools (Rank 1) have higher predicted probabilities of admission compared to lower-ranked schools, confirming that undergraduate rank matters.

### Conclusion ###

GPA was a strong predictor of graduate school admission, with higher GPA values
associated with increased probability of acceptance. GRE scores showed a weaker
effect, suggesting they are less important than GPA.

Undergraduate rank significantly affects admission, with students from top-tier
schools having a higher admission rate. The regression model including GPA, GRE, and
rank (mod3) fits the data best because it has the lowest AIC.

Predicted probabilities and odds ratios provide interpretable insights for
understanding admission likelihood.